---
title: "Mini Project 03 : Visualizing and Maintaining the Green Canopy of NYC"
author: "Lincon Chopra"
engine: knitr
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    theme:
      light: cosmo
      dark: darkly
    df-print: paged
    code-fold: true
execute:
  echo: true
  message: false
  warning: false
  cache: true
---
```{css}
.indent-block-tab {
text-indent: 2em;
}
```


## Introduction
::: {.indent-block-tab}
  In this mini-project, I work with NYC datasets to explore how the city’s green canopy is spread across neighborhoods. The main goal is to build a clean, reproducible process that automatically downloads the spatial data, joins tree locations to council districts and creates clear visualizations of tree coverage.
:::

## Task 1: Download NYC City Council District Boundaries

::: {.indent-block-tab}
  In this task I will write a reusable function that downloads the NYC City Council district data; unzips it if necessary, and transforms the coordinates so it lines up with other spatial data. The function only downloads or unzips the file if it does not already exist in `data/mp03`. The transofrmed data can be veiewed below.
:::


```{r}
library(tidyverse)  
library(sf)         
library(httr2)      
library(glue)
library(janitor)
library(leaflet)

set.seed(9750)

download_council_districts <- function(
  dest_dir = "data/mp03",
  zip_url  = "https://s-media.nyc.gov/agencies/dcp/assets/files/zip/data-tools/bytes/city-council/nycc_25c.zip"
) {
  # create data directory if needed
  if (!dir.exists(dest_dir)) {
    dir.create(dest_dir, recursive = TRUE)
  }

  # paths for zip file and unzipped contents
  zip_path  <- file.path(dest_dir, "nyc_council_districts.zip")
  unzip_dir <- file.path(dest_dir, "nyc_council_districts")

  # download zip only if it does not already exist
  if (!file.exists(zip_path)) {
    download.file(url = zip_url, destfile = zip_path, mode = "wb")
  }

  # check for existing shapefiles in unzip_dir
  if (dir.exists(unzip_dir)) {
    shp_files <- dir(unzip_dir, pattern = "\\.shp$", full.names = TRUE, recursive = TRUE)
  } else {
    shp_files <- character(0)
  }

  # unzip only if we do not already have a shapefile
  if (!dir.exists(unzip_dir) || length(shp_files) == 0) {
    dir.create(unzip_dir, recursive = TRUE, showWarnings = FALSE)
    unzip(zipfile = zip_path, exdir = unzip_dir)
  }

  # find shapefile path
  shp_path <- dir(unzip_dir, pattern = "\\.shp$", full.names = TRUE, recursive = TRUE)[1]

  # read shapefile as sf object
  districts_raw <- sf::st_read(shp_path, quiet = TRUE)

  # transform to WGS84 so it aligns with other data
  districts_wgs84 <- districts_raw |>
    sf::st_transform(crs = "WGS84")

  # create district_id column if needed
  if (!("district_id" %in% names(districts_wgs84))) {
    idx <- which(grepl("coun", names(districts_wgs84), ignore.case = TRUE))[1]
    if (!is.na(idx)) {
      districts_wgs84$district_id <- districts_wgs84[[idx]]
    }
  }

  # ensure district_id is integer
  if ("district_id" %in% names(districts_wgs84)) {
    districts_wgs84$district_id <- as.integer(districts_wgs84$district_id)
  }

  # return sf object
  districts_wgs84
}

# run function to get council_sf
council_sf <- download_council_districts()

# quick check (no geometry)
council_sf |>
  sf::st_drop_geometry() |>
  head()
```
## Task 2: Download Tree Points

::: {.indent-block-tab}
In this task, I write a function that downloads the NYC Forestry Tree Points directly from the NYC OpenData API using the `httr2` package. The function looks through the API and only downloads a file if it does not already exist. 
:::

```{r}
download_tree_points <- function(
  dest_dir    = "data/mp03",
  base_url    = "https://data.cityofnewyork.us/resource/hn5i-inap.geojson",
  chunk_limit = 50000  # number of rows per API request
) {
  # ensure base data directory exists
  if (!dir.exists(dest_dir)) {
    dir.create(dest_dir, recursive = TRUE)
  }

  # directory for raw GeoJSON chunks
  raw_dir <- file.path(dest_dir, "tree_points_raw")
  if (!dir.exists(raw_dir)) {
    dir.create(raw_dir, recursive = TRUE)
  }

  # start at offset 0 and keep track of chunk file paths
  offset     <- 0L
  file_paths <- character()
  total_rows <- 0L

  # loop over API pages using $limit and $offset
  repeat {
    # build filename for this chunk
    file_name <- glue::glue("treepoints_{offset}_{offset + chunk_limit - 1}.geojson")
    file_path <- file.path(raw_dir, file_name)

    # print progress about which chunk we are working on
    message("Chunk starting at offset = ", offset)

    # download this chunk only if file does not already exist
    if (!file.exists(file_path)) {
      message("  Downloading chunk to: ", file_path)
      resp <- httr2::request(base_url) |>
        httr2::req_url_query(`$limit` = chunk_limit,
                             `$offset` = offset) |>
        httr2::req_perform()

      # save raw response body to disk
      writeBin(httr2::resp_body_raw(resp), file_path)
    } else {
      message("  File already exists, skipping download.")
    }

    # read this chunk to see how many rows it contains
    trees_chunk <- sf::st_read(file_path, quiet = TRUE)
    n_rows      <- nrow(trees_chunk)
    total_rows  <- total_rows + n_rows

    message("  Rows in this chunk: ", n_rows, " | Total rows so far: ", total_rows)

    # store file path so we can re-read later
    file_paths <- c(file_paths, file_path)

    # if fewer than chunk_limit rows, we reached the end of the dataset
    if (n_rows < chunk_limit) {
      message("Reached final chunk (n_rows < chunk_limit). Stopping download loop.")
      break
    }

    # otherwise move offset forward to get the next page
    offset <- offset + chunk_limit
  }

  # read all saved chunk files and combine with rbind (keeps sf class)
  message("Reading all saved chunk files and combining...")
  tree_list <- purrr::map(file_paths, function(path) {
    sf::st_read(path, quiet = TRUE)
  })

  trees_sf <- do.call(rbind, tree_list)

  # transform to WGS84 to match council_sf
  message("Transforming tree coordinates to WGS84...")
  trees_sf <- sf::st_transform(trees_sf, crs = "WGS84")

  message("Done. Total rows: ", nrow(trees_sf))

  # return combined sf object
  trees_sf
}

# run function to get trees_sf 
trees_sf <- download_tree_points()

trees_sf |>
  sf::st_drop_geometry() |>
  head()

```
## Task 3: Data Integration and Initial Exploration

::: {.indent-block-tab}
For this task we will be creating a visualization that will inspect the distribution of trees in NYC. This will allow us to quickly analyze how trees are spread across the city and help us identify areas of concentration.
:::
```{r}
trees_sample <- trees_sf |>
  dplyr::slice_sample(n = min(20000, nrow(trees_sf)))

ggplot() +
  # council district boundaries as light outlines
  geom_sf(data = council_sf,
          fill = NA,
          color = "grey60",
          linewidth = 0.4) +
  # sampled tree points as small, semi-transparent dots
  geom_sf(data = trees_sample,
          color = "darkgreen",
          alpha = 0.15,   # more transparent
          size  = 0.3) +  # smaller points
  coord_sf() +
  labs(
    title    = "NYC Tree Points",
    caption  = "Mapping NYC Trees"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_line(linewidth = 0.2, colour = "grey90"),
    panel.background = element_rect(fill = "white", colour = NA)
  )
```
::: {.indent-block-tab}
This map shows that trees are spread across all five boroughs, but the dense parts appear in Brooklyn, Queens, and the northern half of Manhattan. Staten Island and some waterfront or industrial areas look more patchy, suggesting fewer recorded street trees there. Overall, this provides a quick citywide view of where NYC’s trees is heaviest.
:::

## Task 4: District-Level Analysis of Tree Coverage

::: {.indent-block-tab}
Before we can further analyze our NYC trees, we need to join the Tree Points data onto the District boundaries. Once this is done I will be answering a couple of exploratory questions below. 
:::
```{r}
# tag each tree with its council district
trees_with_district <- sf::st_join(
  trees_sf,
  council_sf,
  join = sf::st_intersects
)

# drop geometry for easier summaries
trees_tbl <- trees_with_district |>
  sf::st_drop_geometry()

# pull district area from council shapefile 
district_area <- council_sf |>
  sf::st_drop_geometry() |>
  dplyr::select(district_id, Shape_Area) |>
  dplyr::rename(area_sq_m = Shape_Area)

# summarise trees per district: counts, dead fraction
district_summary <- trees_tbl |>
  dplyr::group_by(district_id) |>
  dplyr::summarise(
    n_trees = dplyr::n(),
    n_dead  = sum(
      stringr::str_detect(stringr::str_to_lower(tpcondition), "dead"),
      na.rm = TRUE
    ),
    frac_dead = n_dead / n_trees,
    .groups   = "drop"
  ) |>
  dplyr::left_join(district_area, by = "district_id")

# convert area to square miles and compute trees per square mile
sqm_per_sqmi <- 1609.34^2  # number of square meters in one square mile

district_summary <- district_summary |>
  dplyr::mutate(
    area_sq_mi        = area_sq_m / sqm_per_sqmi,
    tree_density_sqmi = n_trees / area_sq_mi
  )

# total trees across all districts (for percentages later)
total_trees <- sum(district_summary$n_trees, na.rm = TRUE)

# district with most trees
district_most_trees <- district_summary |>
  dplyr::slice_max(n_trees, n = 1)

district_most_trees_id <- district_most_trees$district_id

# district with highest tree density (now using trees per square mile)
district_highest_density <- district_summary |>
  dplyr::filter(!is.na(tree_density_sqmi)) |>
  dplyr::slice_max(tree_density_sqmi, n = 1)

district_highest_density_id <- district_highest_density$district_id

# district with highest fraction of dead trees
district_highest_dead <- district_summary |>
  dplyr::filter(!is.na(frac_dead)) |>
  dplyr::slice_max(frac_dead, n = 1)

district_highest_dead_id <- district_highest_dead$district_id

# borough column + Manhattan species
trees_tbl <- trees_tbl |>
  dplyr::mutate(
    borough = dplyr::case_when(
      dplyr::between(district_id, 1, 10)  ~ "Manhattan",
      dplyr::between(district_id, 11, 18) ~ "Bronx",
      dplyr::between(district_id, 19, 32) ~ "Queens",
      dplyr::between(district_id, 33, 48) ~ "Brooklyn",
      dplyr::between(district_id, 49, 51) ~ "Staten Island",
      TRUE                                ~ NA_character_
    )
  )

# make a district → borough lookup table
district_borough_map <- trees_tbl |>
  dplyr::distinct(district_id, borough)

# borough for the district with the most trees
district_most_trees_borough <- district_borough_map |>
  dplyr::filter(district_id == district_most_trees_id) |>
  dplyr::pull(borough)

# borough for the district with highest tree density
district_highest_density_borough <- district_borough_map |>
  dplyr::filter(district_id == district_highest_density_id) |>
  dplyr::pull(borough)

# borough for the district with highest fraction of dead trees
district_highest_dead_borough <- district_borough_map |>
  dplyr::filter(district_id == district_highest_dead_id) |>
  dplyr::pull(borough)

manhattan_species_counts <- trees_tbl |>
  dplyr::filter(
    borough == "Manhattan",
    !is.na(genusspecies),
    genusspecies != ""
  ) |>
  dplyr::count(genusspecies, sort = TRUE)

most_common_manhattan    <- manhattan_species_counts$genusspecies[1]
most_common_manhattan_n  <- manhattan_species_counts$n[1]
total_manhattan_trees    <- sum(manhattan_species_counts$n)

# Baruch point + closest tree

new_st_point <- function(lat, lon, ...) {
  sf::st_sfc(sf::st_point(c(lon, lat)), crs = "WGS84")
}

# Newman Vertical Campus coords
baruch_point <- new_st_point(lat = 40.74194, lon = -73.98583)

# compute distance using trees_with_district so we keep district_id
trees_with_distance <- trees_with_district |>
  dplyr::mutate(
    distance_m = as.numeric(sf::st_distance(geometry, baruch_point))
  )

closest_tree <- trees_with_distance |>
  sf::st_drop_geometry() |>
  dplyr::arrange(distance_m) |>
  dplyr::slice(1)

closest_species     <- closest_tree$genusspecies
closest_distance_m  <- closest_tree$distance_m
closest_district_id <- closest_tree$district_id

# borough for the tree closest to Baruch's campus
closest_borough <- district_borough_map |>
  dplyr::filter(district_id == closest_district_id) |>
  dplyr::pull(borough)

```
### Question 1: Which council district has the most trees?

::: {.indent-block-tab}
The council district with the most trees is **District `r district_most_trees_id`** in **`r district_most_trees_borough`**, with about **`r format(district_most_trees$n_trees, big.mark = ",")`** trees. That’s roughly **`r round(100 * district_most_trees$n_trees / total_trees, 1)`%** of all trees. Showing that this borough has the heaviest concentration.
:::

### Question 2: Which council district has the highest density of trees? 

::: {.indent-block-tab}
When I divide the number of trees by the district area in square miles, the highest tree density shows up in **District `r district_highest_density_id`** which is in **`r district_highest_density_borough`**. This borough has about **`r format(district_highest_density$n_trees, big.mark = ",")`** trees over roughly **`r round(district_highest_density$area_sq_mi, 1)`** square miles. This works out to around **`r round(district_highest_density$tree_density_sqmi, 0)`** trees per square mile. We can see that this borough packs them into a smaller space.
:::

### Question 3: Which district has highest fraction of dead trees out of all trees?

::: {.indent-block-tab}
The district with the highest share of dead trees is **District `r district_highest_dead_id`** in **`r district_highest_dead_borough`**. In this borough, **`r district_highest_dead$n_dead`** out of **`r district_highest_dead$n_trees`** trees are labeled as dead, which is about **`r round(100 * district_highest_dead$frac_dead, 1)`%** of its total trees. That suggests this area might be a good target for maintenance and replacement.
:::

### Question 4: What is the most common tree species in Manhattan?

::: {.indent-block-tab}
The most common tree species in Manhattan is **`r most_common_manhattan`**. 
:::

### Question 5: What is the species of the tree closest to Baruch’s campus?
::: {.indent-block-tab}
The closest species of tree is **`r closest_species`** near Baruch's campus.
:::

## Task 5: NYC Parks Proposal

::: {.indent-block-tab}
**Queens** has been left out far too long. Its our trees and we need it now. I am proposing a project which will help fund a new tree program within the borough of Queens in **District 29 Richmond Hill**. This is important because in our analysis we found out that Queens has the most dead trees listed. This is an issue because we need to be actively working on planting new trees once they die. What this project intends to do is to identify where new trees can be planted in District 29 Richmond Hill, Queens. Then it will provide funding to those areas for those trees to increase the green canopy within the borough. 
:::

```{r}
# point roughly in Richmond Hill, Queens

richmond_point <- new_st_point(
  lat = 40.695,   # approximate latitude for Richmond Hill
  lon = -73.829   # approximate longitude for Richmond Hill
)

# find which council district polygon contains this point
idx_richmond <- which(sf::st_contains(council_sf, richmond_point, sparse = FALSE))

# sf row for the Richmond Hill district
richmond_district_sf <- council_sf[idx_richmond, ]

# focus district info
queens_focus_id      <- richmond_district_sf$district_id
queens_focus_borough <- "Queens"

# restrict to Queens districts using your existing district_borough_map
queens_districts <- district_borough_map |>
  dplyr::filter(borough == "Queens")

# Queens-only summary
queens_summary <- district_summary |>
  dplyr::inner_join(queens_districts, by = "district_id")

# metrics for the Richmond Hill district
queens_focus <- queens_summary |>
  dplyr::filter(district_id == queens_focus_id)

queens_n_trees      <- queens_focus$n_trees
queens_n_dead       <- queens_focus$n_dead
queens_frac_dead    <- queens_focus$frac_dead
queens_density_sqmi <- queens_focus$tree_density_sqmi
queens_area_sqmi    <- queens_focus$area_sq_mi

# citywide averages
avg_density_sqmi   <- mean(district_summary$tree_density_sqmi, na.rm = TRUE)
avg_frac_dead_city <- mean(district_summary$frac_dead,         na.rm = TRUE)

# comparison districts in other boroughs (NOT Queens)
comparison_districts <- district_summary |>
  dplyr::left_join(district_borough_map, by = "district_id") |>
  dplyr::filter(
    district_id != queens_focus_id,   # not your Richmond Hill district
    borough != "Queens"               # and not in Queens at all
  ) |>
  dplyr::slice_max(n_dead, n = 3)     # top 3 dead-tree counts in other boroughs

# combined data frame for dead-tree comparison bar chart
compare_dead_df <- dplyr::bind_rows(
  queens_focus |>
    dplyr::mutate(borough = queens_focus_borough, focus = TRUE),
  comparison_districts |>
    dplyr::mutate(focus = FALSE)
) |>
  dplyr::mutate(
    label = glue::glue("D{district_id} ({borough})")
  )

```

::: {.indent-block-tab}
Queens District `r queens_focus_id` Richmond Hill area has about **`r format(queens_n_trees, big.mark = ",")`** recorded trees, of which roughly **`r format(queens_n_dead, big.mark = ",")`** are marked as dead, or about **`r round(100 * queens_frac_dead, 1)`%**. For this program, I propose to remove and replace at least half of the currently dead trees, which is about **`r round(queens_n_dead * 0.5)`** trees. I also want to add **`r round(queens_n_dead * 0.25)`** new plantings on blocks that currently lack trees. In total, this would mean planting roughly **`r round(queens_n_dead * 0.75)`** new trees within this district.

:::

```{r}
## zoomed-in map for Queens focus district

 #boundary geometry for the Richmond Hill district (your Queens focus)
queens_boundary <- council_sf |>
  dplyr::filter(district_id == queens_focus_id)

# trees inside the focus district, tagged by condition
queens_trees <- trees_with_district |>
  dplyr::filter(district_id == queens_focus_id) |>
  dplyr::mutate(
    alive_dead = dplyr::case_when(
      stringr::str_detect(stringr::str_to_lower(tpcondition), "dead") ~ "Dead",
      TRUE ~ "Alive or unknown"
    )
  )

# all dead trees (we care about these for the project)
queens_dead_trees <- queens_trees |>
  dplyr::filter(alive_dead == "Dead")

# all alive/unknown trees, then choose a sample size
queens_alive_all <- queens_trees |>
  dplyr::filter(alive_dead == "Alive or unknown")

n_alive     <- nrow(queens_alive_all)
sample_size <- min(4000L, n_alive)  # cap at 4000 for readability

queens_alive_sample <- queens_alive_all |>
  dplyr::slice_sample(n = sample_size)

# bounding box to keep the view tightly on this district
bbox_q <- sf::st_bbox(queens_boundary)

ggplot() +
  geom_sf(data = queens_boundary,
          fill      = "grey98",
          color     = "black",
          linewidth = 0.6) +
  geom_sf(data  = queens_alive_sample,
          color = "darkseagreen3",
          alpha = 0.4,
          size  = 0.5) +
  geom_sf(data  = queens_dead_trees,
          color = "red3",
          alpha = 0.9,
          size  = 0.8) +
  coord_sf(
    xlim   = c(bbox_q["xmin"], bbox_q["xmax"]),
    ylim   = c(bbox_q["ymin"], bbox_q["ymax"]),
    expand = FALSE
  ) +
  labs(
    title    = glue::glue("Dead Trees in Queens District {queens_focus_id} (Richmond Hill)"),,
    x        = "Longitude",
    y        = "Latitude",
    caption  = "Red points show all trees marked as dead."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_line(colour = "grey90", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold"),
    legend.position  = "none"
  )
```
::: {.indent-block-tab}
For our proposal I am providing the visualization above which shows a concentration of dead trees within District 29 Richmond Hill, Queens. This lets us visually identify where we need to plant the trees in this project. Its our trees and we need it now!
:::


::: {.indent-block-tab}
Queens District `r queens_focus_id`- Richmond Hill, has about **`r format(queens_n_dead, big.mark = ",")`** dead trees out of **`r format(queens_n_trees, big.mark = ",")`** total, so roughly **`r round(100 * queens_frac_dead, 1)`%** of the canopy is already classified as dead. To put this in context, I compared it to three other high-need districts in other boroughs: **District `r comparison_districts$district_id[1]` (`r comparison_districts$borough[1]`)**, **District `r comparison_districts$district_id[2]` (`r comparison_districts$borough[2]`)**, and **District `r comparison_districts$district_id[3]` (`r comparison_districts$borough[3]`)**. These districts have around **`r format(comparison_districts$n_dead[1], big.mark = ",")`**, **`r format(comparison_districts$n_dead[2], big.mark = ",")`**, and **`r format(comparison_districts$n_dead[3], big.mark = ",")`** dead trees respectively. When comparing to these districts Queens District `r queens_focus_id` Richmond Hill has a similarly high share of its trees in poor condition.
:::

::: {.indent-block-tab}
Looking at canopy density tells a similar story. Queens District `r queens_focus_id` Richmond has about **`r round(queens_density_sqmi, 0)`** trees per square mile, which is close to the citywide average of roughly **`r round(avg_density_sqmi, 0)`** trees per square mile. That means it does not have an surplus to begin with, yet it still has a substantial amount of dead trees. The comparison districts in Staten Island tend to be more heavier in tree density, so losing trees there is buffered by a denser population of trees, whereas in Richmond Hill each dead tree represents a bigger loss relative to the total. 
:::

```{r}
## Bar chart: Queens focus district vs 3 high-dead-tree districts in other boroughs

ggplot(compare_dead_df,
       aes(x = reorder(label, n_dead),
           y = n_dead,
           fill = focus)) +
  # one bar per district
  geom_col() +
  # flip so labels are easier to read
  coord_flip() +
  # highlight Queens focus district, grey for others
  scale_fill_manual(
    values = c("TRUE" = "firebrick", "FALSE" = "grey70"),
    labels = c("TRUE" = "Queens focus district",
               "FALSE" = "Other districts"),
    name   = NULL
  ) +
  labs(
    title    = "Dead trees by NYC council district",
    subtitle = glue::glue("Queens District {queens_focus_id} (Richmond Hill) vs 3 other high-dead-tree districts"),
    x        = "District",
    y        = "Number of trees marked dead"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom"
  )
```
::: {.indent-block-tab}
In this chart it can be seen that even though the other districts hav higher dead tree counts. The density is higher causing a higher buffer point.  
:::
```{r}
## Task 5: map-based comparison of tree density
## Choropleth: trees per square mile for all districts

# attach summary metrics back onto the council geometry
district_map_sf <- council_sf |>
  dplyr::left_join(district_summary, by = "district_id")

ggplot(district_map_sf) +
  # fill each district by tree density per square mile
  geom_sf(aes(fill = tree_density_sqmi),
          color     = "white",
          linewidth = 0.2) +
  # color scale with sqrt transform so extremes are easier to see
  scale_fill_viridis_c(
    option = "viridis",
    trans  = "sqrt",
    name   = "Trees per\nsquare mile"
  ) +
  labs(
    title    = "Tree density by NYC City Council District"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.background = element_rect(fill = "grey98", colour = NA),
    panel.grid.major = element_line(colour = "grey90", linewidth = 0.2)
  )

```
::: {.indent-block-tab}
In this map based visual we can clearly see that our district within the borough of queens has a lower trees per square mile ratio. This further supports my proposal of implementing tree revival within the district 29 Richmond Hill Queens. Its our trees and we need it now!!!  
:::

## Extra Credit Opportunity #01: Improved Tree Map Visualizations
```{r}


library(leaflet)

# make sure we have the Richmond Hill district sf object
richmond_district_sf <- council_sf |>
  dplyr::filter(district_id == queens_focus_id)

 

# tag each tree as Dead vs Alive
trees_for_map <- trees_sf |>
  dplyr::mutate(
    cond_group = dplyr::case_when(
      stringr::str_detect(stringr::str_to_lower(tpcondition), "dead") ~ "Dead",
      TRUE ~ "Alive"
    )
  )

# split into dead and alive, then sample for performance
dead_all <- trees_for_map |>
  dplyr::filter(cond_group == "Dead")

alive_all <- trees_for_map |>
  dplyr::filter(cond_group == "Alive")

dead_sample_size  <- min(15000L, nrow(dead_all))
alive_sample_size <- min(15000L, nrow(alive_all))

dead_sample <- dead_all |>
  dplyr::slice_sample(n = dead_sample_size)

alive_sample <- alive_all |>
  dplyr::slice_sample(n = alive_sample_size)

# convert sf objects to plain data frames with lon/lat columns
coords_dead  <- sf::st_coordinates(dead_sample)
coords_alive <- sf::st_coordinates(alive_sample)

dead_df <- dead_sample |>
  sf::st_drop_geometry() |>
  dplyr::mutate(
    lon = coords_dead[, 1],
    lat = coords_dead[, 2]
  )

alive_df <- alive_sample |>
  sf::st_drop_geometry() |>
  dplyr::mutate(
    lon = coords_alive[, 1],
    lat = coords_alive[, 2]
  )



district_centers <- sf::st_centroid(council_sf)

coords_centers <- sf::st_coordinates(district_centers)

district_centers <- district_centers |>
  dplyr::mutate(
    lon = coords_centers[, 1],
    lat = coords_centers[, 2]
  )



leaflet() |>
  # base map tiles
  addProviderTiles(providers$CartoDB.Positron) |>
  
  # all council districts outlined in black
  addPolygons(
    data        = council_sf,
    fillOpacity = 0,
    color       = "black",
    weight      = 1,
    label       = ~paste("District", district_id)
  ) |>
  
  # Richmond Hill focus district highlighted in red
  addPolygons(
    data        = richmond_district_sf,
    fillOpacity = 0,
    color       = "red",
    weight      = 3
  ) |>
  
  # district numbers at centroids
  addLabelOnlyMarkers(
    data  = district_centers,
    lng   = ~lon,
    lat   = ~lat,
    label = ~as.character(district_id),
    labelOptions = labelOptions(
      noHide    = TRUE,
      direction = "center",
      textsize  = "9px",
      opacity   = 0.7
    )
  ) |>
  
  # alive trees layer (green)
  addCircleMarkers(
    data        = alive_df,
    lng         = ~lon,
    lat         = ~lat,
    radius      = 2,
    stroke      = FALSE,
    fillOpacity = 0.4,
    color       = "darkgreen",
    group       = "Alive"
  ) |>
  
  # dead trees layer (red)
  addCircleMarkers(
    data        = dead_df,
    lng         = ~lon,
    lat         = ~lat,
    radius      = 3,
    stroke      = FALSE,
    fillOpacity = 0.7,
    color       = "red",
    group       = "Dead"
  ) |>
  
  # layer control to toggle alive/dead trees
  addLayersControl(
    overlayGroups = c("Alive", "Dead"),
    options       = layersControlOptions(collapsed = FALSE)
  ) |>
  
  # legend for tree status
  addLegend(
    position = "bottomright",
    colors   = c("darkgreen", "red"),
    labels   = c("Alive", "Dead"),
    title    = "Tree status"
  )

```

