---
title: "Mini-Project 04: Just the Fact(-Check)s, Ma’am!"
author: "Lincon Chopra"
engine: knitr
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    theme:
      light: cosmo
      dark: darkly
    df-print: paged
    code-fold: true
execute:
  echo: true
  message: false
  warning: false
  cache: true
---
```{css}
.indent-block-tab {
text-indent: 2em;
}
```

## Introduction
::: {.indent-block-tab}
In this mini-project, I will be looking into President Trump’s 2025 decision to fire BLS Commissioner Dr. Erika McEntarfer over controversies about revisions to monthly job numbers. Using data tools we’ve learned, I will fact check a public claim about the accuracy of BLS reports.
:::

## Task 1: Download CES Total Nonfarm Payroll
::: {.indent-block-tab}
In this task I will be using httr2 and rvest to access and download the CES Total Nonfarm Payroll Seasonally Adjusted for each month from January 1979 to June of 2025.
:::
```{r}
library(httr2)
library(rvest)
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(purrr)
library(ggplot2)
library(infer)
```


```{r}

## 1. Request the CES total nonfarm series table (all years, all months)
resp <- request("https://data.bls.gov/pdq/SurveyOutputServlet") |>
  req_method("POST") |>
  req_body_form(
    survey    = "ce",
    series_id = "CES0000000001",  # total nonfarm, seasonally adjusted
    from_year = "1979",
    to_year   = "2025"
  ) |>
  req_perform()

## 2. Parse HTML from the response
ces_html <- resp |> resp_body_html()

## 3. Grab the main data table
ces_tables <- ces_html |>
  html_elements("table") |>
  html_table(fill = TRUE)

# Pick the table that has a 'Year' column
ces_raw <- ces_tables |>
  keep(~ "Year" %in% names(.x)) |>
  pluck(1)

## 4. Clean and reshape to date / level
ces_tidy <-
  ces_raw |>
  # keep only rows that are actual years (drop headers/footers)
  filter(str_detect(Year, "^[0-9]{4}$")) |>
  mutate(Year = as.integer(Year)) |>
  pivot_longer(
    cols      = -Year,
    names_to  = "month",
    values_to = "level"
  ) |>
  # keep only real months, drop "Annual" or other summary columns
  filter(month %in% month.abb) |>
  mutate(
    # remove footnote chars (like * or p), keep digits, dot, minus
    level = as.numeric(gsub("[^0-9.-]", "", level)),
    # build a proper Date from "1979 Jan" → 1979-01-01
    date  = ym(paste(Year, month))
  ) |>
  arrange(date) |>
  # restrict to Jan 1979 – Jun 2025
  filter(
    date >= ymd("1979-01-01"),
    date <= ymd("2025-06-01")
  ) |>
  select(date, level) |>
  drop_na()

ces_tidy


```

## Task 2: Download CES Revisions Tables
::: {.indent-block-tab}
Here I use httr2 with a custom User Agent to download the CES revisions HTML page from BLS; then use rvest to scrape the tables directly from that page.
:::
```{r}

## 1. Get the revisions page with httr2 (BLS blocks "generic" requests)
rev_resp <- request("https://www.bls.gov") |>
  req_url_path("web", "empsit", "cesnaicsrev.htm") |>
  req_headers(
    `User-Agent` = "Mozilla/5.0 (Macintosh; Intel Mac OS X 10.15; rv:143.0) Gecko/20100101 Firefox/143.0"
  ) |>
  req_error(is_error = \(resp) FALSE) |>
  req_perform()

resp_check_status(rev_resp)

rev_html <- resp_body_html(rev_resp)

## 2. Grab ALL tables on the page
rev_tables <- rev_html |> html_elements("table")

## 3. Helper: pull out the month/year + 1st & 3rd estimates from ONE table
extract_revision_rows <- function(tbl_node) {
  df <- tbl_node |>
    html_table(header = FALSE, fill = TRUE) |>
    as_tibble()

  # Give generic names V1, V2, ...
  names(df) <- paste0("V", seq_len(ncol(df)))

  # Need at least 5 columns: month, year, 1st, 2nd, 3rd
  if (ncol(df) < 5) return(NULL)

  # Work with character for month/year
  df <- df |>
    mutate(
      V1 = as.character(V1),
      V2 = as.character(V2)
    )

  # Match "Jan.", "Feb.", ..., with or without the dot
  month_pat <- "^(Jan|Feb|Mar|Apr|May|Jun|Jul|Aug|Sep|Oct|Nov|Dec)\\.?$"

  # Keep rows that look like a real month + 4-digit year
  rows <- df |>
    filter(
      str_detect(V1, month_pat),
      str_detect(V2, "^[0-9]{4}$")
    )

  if (nrow(rows) == 0) return(NULL)

  rows |>
    transmute(
      year     = as.integer(V2),
      # strip trailing "." so lubridate::ym can handle it
      month    = str_remove(V1, "\\.$"),
      # Seasonally adjusted 1st and 3rd estimates = original & final
      original = as.numeric(gsub("[^0-9-]", "", V3)),
      final    = as.numeric(gsub("[^0-9-]", "", V5))
    )
}

## 4. Apply to all tables and stack them
revisions_raw <- rev_tables |>
  map(extract_revision_rows) |>
  list_rbind()

## 5. Build final dataset: date, original, final, revision
ces_revisions <- revisions_raw |>
  mutate(
    date     = ym(paste(year, month)),
    revision = final - original
  ) |>
  # Keep only complete numeric rows in desired range
  filter(
    !is.na(date),
    !is.na(original),
    !is.na(final),
    date >= ymd("1979-01-01"),
    date <= ymd("2025-06-01")
  ) |>
  arrange(date) |>
  select(date, original, final, revision)

ces_revisions
```

## Task 3: Data Exploration and Visualization
::: {.indent-block-tab}
In task 3, I will be joining together the above tables and perform exploratory data analysis to investigate trends in CES levels and revisions.
:::

```{r}
# 1) Join the CES levels and revision tables on 'date'
# 2) Create some helper variables for analysis (year, month, etc.)
# 3) Compute absolute and relative revision measures
ces_full <- ces_tidy |>
  left_join(ces_revisions, by = "date") |>
  mutate(
    # Time helpers
    year       = year(date),
    month      = month(date),
    month_name = month(date, label = TRUE, abbr = TRUE),

    # Revision magnitude in absolute terms
    abs_revision = abs(revision),

    # Revision as a fraction of the overall level (from Task 1)
    revision_pct_level     = revision / level,
    abs_revision_pct_level = abs(revision) / level,

    # Revision as a fraction of the final (Task 2)
    revision_pct_final     = revision / final,
    abs_revision_pct_final = abs(revision) / final,

    # Group years into decades (e.g. 1982 -> 1980)
    decade = floor(year / 10) * 10
  )

ces_full
```

::: {.indent-block-tab}
Here are some statistics about the CES over the past 45 years. These stats are all calculated from Jan 1979 through Jun 2025, so `n_months` is just the number of months where we actually have a revision. The table shows whether revisions tend to be up or down on average (`mean_revision`), how big they usually are in both raw job counts and as a share of total employment (`mean_abs_revision`, `mean_revision_pct_lvl`, `mean_abs_rev_pct_lvl`), how often BLS revises the number upward (`frac_positive_rev`), and the biggest upward and downward monthly revisions (`max_revision`, `min_revision`).
:::

```{r}
# Overall summary statistics about revisions
ces_summary <- ces_full |>
  summarise(
    n_months              = sum(!is.na(revision)),           # how many months with revision data
    mean_revision         = mean(revision, na.rm = TRUE),    # average signed revision
    mean_abs_revision     = mean(abs_revision, na.rm = TRUE),# average absolute revision
    mean_revision_pct_lvl = mean(revision_pct_level, na.rm = TRUE),
    mean_abs_rev_pct_lvl  = mean(abs_revision_pct_level, na.rm = TRUE),
    frac_positive_rev     = mean(revision > 0, na.rm = TRUE),# share of months revised up
    max_revision          = max(revision, na.rm = TRUE),     # largest upward revision
    min_revision          = min(revision, na.rm = TRUE)      # largest downward revision
  )

ces_summary

```

::: {.indent-block-tab}
I will now show the months with the most revisions. The first will show the positive and the second the negative. From here we can identify that **11/2021** had the most positive revision and **03/2020** had the most negative revision probably due to covid. 
:::

```{r}
# Month with largest positive revision
largest_pos_rev <- ces_full |>
  filter(!is.na(revision)) |>
  slice_max(revision, n = 1, with_ties = FALSE) |>
  select(date, level, original, final, revision)

# Month with largest negative revision
largest_neg_rev <- ces_full |>
  filter(!is.na(revision)) |>
  slice_min(revision, n = 1, with_ties = FALSE) |>
  select(date, level, original, final, revision)

largest_pos_rev
largest_neg_rev

```
::: {.indent-block-tab}
For a given year or decade we created the following analysis to see if revisions were generally positive or negative. The first table showing the year view and the second showing the decade view. 
:::
```{r}
# Fraction of positive revisions by year
pos_by_year <- ces_full |>
  filter(!is.na(revision)) |>
  group_by(year) |>
  summarise(
    frac_positive = mean(revision > 0),
    mean_abs_rev_pct_level = mean(abs_revision_pct_level, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# Fraction of positive revisions by decade
pos_by_decade <- ces_full |>
  filter(!is.na(revision)) |>
  group_by(decade) |>
  summarise(
    frac_positive = mean(revision > 0),
    mean_abs_rev_pct_level = mean(abs_revision_pct_level, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

pos_by_year
pos_by_decade
```

::: {.indent-block-tab}
Now we will construct 4 visualizations of CES estimates and accuracy over the past 45 years
:::

```{r}
#| label: fig-level
#| fig-cap: "Employment level over time: 1979–2025"

ggplot(ces_full, aes(x = date, y = level / 1000)) +
geom_line() +
labs(
title = "Employment level over time: 1979–2025",
x = NULL,
y = "Employment (thousands)"
)

```

::: {.indent-block-tab}
With this snapshot we are able to clearly see upward and downward trends over time in terms of employment numbers. 
:::
```{r}
#| label: fig-revisions-ts
#| fig-cap: "CES revisions over time (final – original)."

ggplot(ces_full, aes(x = date, y = revision)) +
geom_hline(yintercept = 0, linetype = "dashed") +
geom_line(alpha = 0.7) +
labs(
title = "CES Revisions Over Time",
x = NULL,
y = "Revision (thousands)"
)

```
::: {.indent-block-tab}
With this view we are able to identify if revisions bounce between positive and negative over time.
:::
```{r}
#| label: fig-revisions-by-month
#| fig-cap: "Distribution of CES revisions by month of year."

ggplot(ces_full, aes(x = month_name, y = revision)) +
geom_boxplot() +
labs(
title = "Distribution of CES Revisions by Month of Year",
x = "Month",
y = "Revision"
)

```
::: {.indent-block-tab}
It can be seen that the later months had higher positive outliers versus the beginning months within the year. 
:::
```{r}
#| label: fig-abs-rev-pct
#| fig-cap: "Absolute CES revision as % of employment level."

ggplot(ces_full, aes(x = date, y = abs_revision_pct_level * 100)) +
geom_line(alpha = 0.7) +
labs(
title = "Absolute CES Revision as % of Employment Level",
x = NULL,
y = "Absolute revision %)"
)

```
::: {.indent-block-tab}
Now we are able to see if revisions are becoming relatively more or less important over time taking into account that the job market today is significantly larger than the past. 
:::

## Task 4: Statistical Inference
::: {.indent-block-tab}
For this task I will be using t_test and prop_test functions to perform statistical tests about CES revisions. From the one sample t-test, on average, the CES estimate understates employment by around **12k jobs**, and the final numbers are revised up. So there is evidence of a small, systematic upward revision over the full period. When looking at the two sample t-test pre and post 2020 we don’t see strong evidence that average revisions changed in a meaningful way after 2020.Lastly based on the prop-test there’s no strong evidence that negative revisions became more or less common after 2000.
:::
```{r}
# helper columns
ces_full <- ces_full |>
  mutate(
    year               = year(date),
    abs_revision       = abs(revision),
    revision_pct_level = revision / level
  )


# Test 1: Is the average revision different from zero?
#   H0: mean(revision) = 0
#   HA: mean(revision) ≠ 0
#   (one-sample t-test)

rev_ttest_overall <- ces_full |>
  filter(!is.na(revision)) |>
  t_test(
    response = revision,  # numeric response
    mu       = 0          # null hypothesis mean
  )

cat("\n Test 1: One-sample t-test for mean revision (overall) \n")
rev_ttest_overall
# estimate  = sample mean revision
# p_value   = evidence against "average revision is 0"



# Test 2: Has the average revision changed post-2020?
#   Compare mean revision for pre-2020 vs post-2020
#   H0: mean_post2020 - mean_pre2020 = 0
#   HA: ≠ 0 (two-sided)
#   (two-sample t-test)

rev_ttest_2020 <- ces_full |>
  filter(!is.na(revision)) |>
  mutate(
    era_2020 = if_else(date >= ymd("2020-01-01"), "post_2020", "pre_2020")
  ) |>
  t_test(
    revision ~ era_2020,                 # numeric ~ group
    order = c("post_2020", "pre_2020"),  # estimate = post - pre
    alternative = "two.sided"
  )

cat("\n Test 2: Two-sample t-test for mean revision (post-2020 vs pre-2020) \n")
rev_ttest_2020
# estimate  = mean revision (post-2020) - mean revision (pre-2020)
# p_value   = evidence that the mean revision changed after 2020



# Test 3: Has the fraction of NEGATIVE revisions increased post-2000?
#   Response: "negative" vs "non_negative" (categorical)
#   Group:    pre_2000 vs post_2000
#   H0: p_post2000 - p_pre2000 = 0
#   HA: ≠ 0 (two-sided)
#   (two-sample proportion test)

neg_prop_test_2000 <- ces_full |>
  filter(!is.na(revision)) |>
  mutate(
    # categorical response for prop_test()
    negative_cat = if_else(revision < 0, "negative", "non_negative"),
    # group variable
    era_2000     = if_else(year >= 2000, "post_2000", "pre_2000")
  ) |>
  prop_test(
    negative_cat ~ era_2000,
    order   = c("post_2000", "pre_2000"),  # estimate = post - pre
    success = "negative"                   # which level counts as a "success"
  )

cat("\nTest 3: Proportion test for negative revisions (post-2000 vs pre-2000) \n")
neg_prop_test_2000
# estimate  = p(negative | post-2000) - p(negative | pre-2000)
# p_value   = evidence the fraction of negative revisions changed

```

## Task 5: Fact Checks of Claims about BLS
::: {.indent-block-tab}
  Based on all of the data we have gathered and analyzed, I will now be fact checking two claims about the BLS. The first claim is by President Trump. In his post announcing Erika McEntarfer’s firing, Trump said the jobs report was rigged to make him look bad and pointed to large downward revisions.
  
Between Jan 1979–Jun 2025, there are `r ces_summary$n_months` months with non-missing revisions. The **average revision** is about `r round(ces_summary$mean_revision, 1)` thousand jobs, meaning the final estimate is slightly higher than the initial one, not lower. The **average absolute revision** is roughly `r round(ces_summary$mean_abs_revision, 0)` thousand jobs, which is only about `r round(ces_summary$mean_abs_rev_pct_lvl * 100, 2)`% of the total employment level. As shown in @fig-revisions-ts, revisions bounce above and below zero throughout the entire sample rather than being mostly negative in recent years. @fig-revisions-by-month similarly shows a mix of positive and negative revisions in every month of the year.

We also did some hypothesis testing to see if the claims are supported or denied. 

1. Estimated mean revision: `r round(rev_ttest_overall$estimate, 1)` thousand jobs  
2. 95% CI: [`r round(rev_ttest_overall$lower_ci, 1)`, `r round(rev_ttest_overall$upper_ci, 1)`]  
3. p-value: `r signif(rev_ttest_overall$p_value, 3)`

Because the confidence interval is entirely positive and the p-value is extremely small I am going to reject this claim. Instead, the data suggest that revisions are **on average upward**, which contradicts the idea that BLS is rigging the numbers to make conditions look worse.Therefore we will be rating this claim as **Pants on Fire**.

The second claim we will be looking into would be by Bobby Scott from Virginia. After McEntarfer’s firing, Scott said that The Bureau of Labor Statistics has always served as a nonpartisan, data-driven federal agency that provides key information vital to the United States economy. Between January 1979 through June 2025, there are `r ces_summary$n_months` months with non-missing revision data. The **average absolute revision** is about `r round(ces_summary$mean_abs_revision, 0)` thousand jobs, while the typical employment level is on the ord. On average, revisions are therefore only about `r round(ces_summary$mean_abs_rev_pct_lvl * 100, 2)`% of the employment level, far below the error rate some commentators have claimed. 

@fig-abs-rev-pct shows the **absolute CES revision as a percentage of the employment level** over time. There are a few spikes around major shocks like recessions and COVID, but the baseline stays quite low and there is no long stretch where revisions are consistently huge. Taken together with @fig-level, which shows a steadily growing labor market with occasional downturns, this suggests that revision behavior is pretty stable relative to the scale of the economy. 

To check whether recent years looked off or not, I compared the average revision **pre-2020 vs post-2020** using a two-sample t-test. The estimated difference in mean revision is `r round(rev_ttest_2020$estimate, 1)` thousand jobs, with a 95% confidence interval of [`r round(rev_ttest_2020$lower_ci, 1)`, `r round(rev_ttest_2020$upper_ci, 1)`] thousand jobs and a p-value of `r signif(rev_ttest_2020$p_value, 3)`. Because the confidence interval includes zero and the p-value is large, I reject that average revisions in the 2020s are systematically different from the earlier period. Therefore I will rate the claim as **Mostly True**.

:::


